---
title: "Exploring Data Frames"
teaching: 20
exercises: 10
questions:
- "How can I manipulate a data frame?"
objectives:
- "Remove rows with `NA` values."
- "Append two data frames."
- "Understand what a `factor` is."
- "Convert a `factor` to a `character` vector and vice versa."
- "Display basic properties of data frames including size and class of the columns, names, and first few rows."
keypoints:
- "Use `cbind()` to add a new column to a data frame."
- "Use `rbind()` to add a new row to a data frame."
- "Remove rows from a data frame."
- "Use `na.omit()` to remove rows from a data frame with `NA` values."
- "Use `levels()` and `as.character()` to explore and manipulate factors."
- "Use `str()`, `nrow()`, `ncol()`, `dim()`, `colnames()`, `rownames()`, `head()`, and `typeof()` to understand the structure of a data frame."
- "Read in a csv file using `read.csv()`."
- "Understand what `length()` of a data frame represents."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
```

At this point, you've seen it all: in the last lesson, we toured all the basic
data types and data structures in R. Everything you do will be a manipulation of
those tools. But most of the time, the star of the show is the data frame—the table that we created by loading information from a csv file. In this lesson, we'll learn a few more things
about working with data frames.

## Realistic example

We already learned that the columns of a data frame are vectors, so that our
data are consistent in type throughout the columns. 
So far, you have seen the basics of manipulating data frames with our nordic data; now let’s use those skills to digest a more extensive dataset. Let’s read in the gapminder dataset that we downloaded previously:

```{r, echo = FALSE}
gapminder <- read.csv("data/gapminder_data.csv")
```

> ## Miscellaneous Tips
>
> * Another type of file you might encounter are tab-separated value files
>   (.tsv). To specify a tab as a separator, use `"\\t"` or `read.delim()`.
>
> * Files can also be downloaded directly from the Internet into a local folder
> of your choice onto your computer using the `download.file` function. The
> `read.csv` function can then be executed to read the downloaded file from the
> download location, for example,
>
> ```{r eval=FALSE, echo=TRUE}
> download.file("https://raw.githubusercontent.com/datacarpentry/r-intro-geospatial/master/_episodes_rmd/data/gapminder_data.csv",
>               destfile = "data/gapminder_data.csv")
> gapminder <- read.csv("data/gapminder_data.csv")
> ```
>
> * Alternatively, you can also read in files directly into R from the Internet
>   by replacing the file paths with a web address in `read.csv`. One should
>   note that in doing this no local copy of the csv file is first saved onto
>   your computer. For example,
> 
> ```{r eval=FALSE, echo=TRUE}
> gapminder <- read.csv("https://raw.githubusercontent.com/datacarpentry/r-intro-geospatial/master/_episodes_rmd/data/gapminder_data.csv")
> ```
>
> * You can read directly from excel spreadsheets without
> converting them to plain text first by using the [readxl](https://cran.r-project.org/web/packages/readxl/index.html) package.
{: .callout}

Let's investigate the `gapminder` data frame a bit; the first thing we should
always do is check out what the data looks like with `str`:

```{r}
str(gapminder)
```

We can also examine individual columns of the data frame with our `class` function:

```{r}
class(gapminder$year)
class(gapminder$country)
str(gapminder$country)
```

We can also interrogate the data frame for information about its dimensions;
remembering that `str(gapminder)` said there were 1704 observations of 6
variables in gapminder, what do you think the following will produce, and why?

```{r, results="hide"}
length(gapminder)
```

A fair guess would have been to say that the length of a data frame would be the
number of rows it has (1704), but this is not the case; it gives us the number of columns.

```{r}
class(gapminder)
```

To get the number of rows and columns in our dataset, try:

```{r}
nrow(gapminder)
ncol(gapminder)
```

Or, both at once:

```{r}
dim(gapminder)
```

We'll also likely want to know what the titles of all the columns are, so we can
ask for them later:

```{r}
colnames(gapminder)
```

At this stage, it's important to ask ourselves if the structure R is reporting
matches our intuition or expectations; do the basic data types reported for each
column make sense? If not, we need to sort any problems out now before they turn
into bad surprises down the road, using what we've learned about how R
interprets data, and the importance of *strict consistency* in how we record our
data.

Once we're happy that the data types and structures seem reasonable, it's time
to start digging into our data proper. Check out the first few lines:

```{r}
head(gapminder)
```

> ## Challenge 1
>
> It's good practice to also check the last few lines of your data and some in
> the middle. How would you do this?
>
> Searching for ones specifically in the middle isn't too hard but we could
> simply ask for a few lines at random. How would you code this?
>
> > ## Solution to Challenge 1
> >
> > To check the last few lines it's relatively simple as R already has a
> > function for this:
> > 
> > ```{r, results='hide'}
> > tail(gapminder)
> > tail(gapminder, n = 15)
> > ```
> > 
> > What about a few arbitrary rows just for sanity (or insanity depending on
> > your view)?
> >
> > There are several ways to achieve this.
> >
> > The solution here presents one form using nested functions. i.e. a function
> > passed as an argument to another function. This might sound like a new
> > concept but you are already using it in fact.
> >
> > Remember `my_dataframe[rows, cols]` will print to screen your data frame
> > with the number of rows and columns you asked for (although you might 
> > have asked for a range or named columns for example). How would you get the
> > last row if you don't know how many rows your data frame has? R has a
> > function for this. What about getting a (pseudorandom) sample? R also has a
> > function for this.
> > 
> > ```{r, results="hide"}
> > gapminder[sample(nrow(gapminder), 5), ]
> > ```
> {: .solution}
{: .challenge}

> ## Challenge 2
>
> Read the output of `str(gapminder)` again; this time, use what you've learned
> about factors and vectors, as well as the output of functions like `colnames`
> and `dim` to explain what everything that `str` prints out for `gapminder`
> means. If there are any parts you can't interpret, discuss with your
> neighbors!
>
> > ## Solution to Challenge 2
> >
> > The object `gapminder` is a data frame with columns
> >
> > - `country` and `continent` are factors.
> > - `year` is an integer vector.
> > - `pop`, `lifeExp`, and `gdpPercap` are numeric vectors.
> >
> {: .solution}
{: .challenge}

## Adding columns and rows in data frames

We would like to create a new column to hold information on whether the life expectancy is below the world average life expectancy (70.5) or above:

```{r}
below_average <- gapminder$lifeExp < 70.5
head(gapminder)
```

We can then add this as a column via:

```{r, eval=FALSE}
cbind(gapminder, below_average)
```

```{r, eval=TRUE, echo=FALSE}
head(cbind(gapminder, below_average))
```

We probably don't want to print the entire dataframe each time, so
let's put our `cbind` command within a call to `head` to return
only the first six lines of the output.

```{r, eval=TRUE}
head(cbind(gapminder, below_average))
```

Note that if we tried to add a vector of `below_average` with a different number of entries than the number of rows in the dataframe, it would fail:

```{r, error=TRUE}
below_average <- c(TRUE, TRUE, TRUE, TRUE, TRUE)
head(cbind(gapminder, below_average))
```

Why didn't this work? R wants to see one element in our new column
for every row in the table:

```{r}
nrow(gapminder)
length(below_average)
```

So for it to work we need either to have `nrow(gapminder)` = `length(below_average)` or `nrow(gapminder)` to be a multiple of `length(below_average)`: 

```{r, error=TRUE}
below_average <- c(TRUE, TRUE, FALSE)
head(cbind(gapminder, below_average))
```
The sequence `TRUE,TRUE,FALSE` is repeated over all the gapminder rows.

Let's overwite the content of gapminder with our new data frame.

```{r}
below_average <-  as.logical(gapminder$lifeExp<70.5)
gapminder <- cbind(gapminder, below_average)
```

Now how about adding rows? The rows of a data frame are lists:

```{r}
new_row <- list('Norway', 2016, 5000000, 'Nordic', 80.3, 49400.0, FALSE)
gapminder_norway <- rbind(gapminder, new_row)
tail(gapminder_norway)
```

To understand why R is giving us a warning when we try to add this row, let's learn a little more about factors.

## Factors

Here is another thing to look out for: in a `factor`, each different value
represents what is called a `level`. In our case, the `factor` "continent" has 5
levels: "Africa", "Americas", "Asia", "Europe" and "Oceania". R will only accept
values that match one of the levels. If you add a new value, it will become
`NA`.

The warning is telling us that we unsuccessfully added "Nordic" to our
*continent* factor, but 2016 (a numeric), 5000000 (a numeric), 80.3 (a numeric),
49400.0 (a numeric) and `FALSE` (a logical) were successfully added to
*country*, *year*, *pop*, *lifeExp*, *gdpPercap* and *below_average*
respectively, since those variables are not factors. 'Norway' was also
successfully added since it corresponds to an existing level. To successfully
add a gapminder row with a "Nordic" *continent*, add "Nordic" as a *level* of
the factor:

```{r}
levels(gapminder$continent)
levels(gapminder$continent) <- c(levels(gapminder$continent), "Nordic")
gapminder_norway  <- rbind(gapminder,
                           list('Norway', 2016, 5000000, 'Nordic', 80.3,49400.0, FALSE))
tail(gapminder_norway)
```

Alternatively, we can change a factor into a character vector; we lose the handy
categories of the factor, but we can subsequently add any word we want to the
column without babysitting the factor levels:

```{r}
str(gapminder)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

## Appending to a data frame

The key to remember when adding data to a data frame is that *columns are
vectors and rows are lists.* We can also glue two data frames together with
`rbind`:

```{r}
gapminder <- rbind(gapminder, gapminder)
tail(gapminder, n=3)
```
But now the row names are unnecessarily complicated (not consecutive numbers).
We can remove the rownames, and R will automatically re-name them sequentially:

```{r}
rownames(gapminder) <- NULL
head(gapminder)
```

> ## Challenge 3
>
> You can create a new data frame right from within R with the following syntax:
> ```{r}
> df <- data.frame(id = c("a", "b", "c"),
>                  x = 1:3,
>                  y = c(TRUE, TRUE, FALSE),
>                  stringsAsFactors = FALSE)
> ```
>
> Make a data frame that holds the following information for yourself:
>
> - first name
> - last name
> - lucky number
>
> Then use `rbind` to add an entry for the people sitting beside you. Finally,
> use `cbind` to add a column with each person's answer to the question, "Is it
> time for coffee break?"
>
> > ## Solution to Challenge 3
> >
> > ```{r}
> > df <- data.frame(first = c("Grace"),
> >                  last = c("Hopper"),
> >                  lucky_number = c(0),
> >                  stringsAsFactors = FALSE)
> > df <- rbind(df, list("Marie", "Curie", 238) )
> > df <- cbind(df, coffeetime = c(TRUE, TRUE))
> > ```
> {: .solution}
{: .challenge}

